From Reversible Programs to \\ Univalent Universes and Back

### 2017-05-03 upi.tex
entcs style...

\documentclass[format=acmlarge,review,natbib]{acmart}

@@

\title{From Reversible Programs to Univalent Universes and Back}
\author{Jacques Carette}
\affiliation{
  \institution{McMaster University}
  \country{Canada}}
  \email{carette@mcmaster.ca}
\author{Chao-Hong Chen}
\author{Vikraman Choudhury}
\author{Robert Rose}
\author{Amr Sabry}
\affiliation{
  \institution{Indiana University}
  \country{United States}}
\email{\{chen464,vikraman,rrose1,sabry\}@indiana.edu}

@@

\maketitle

@@

The practice of programming languages is replete with \emph{ad hoc} instances of
reversible computations such as database transactions, mechanisms for data
provenance, checkpoints, stack and exception traces, logs, backups, rollback
recoveries, version control systems, reverse engineering, software transactional
memories, continuations, backtracking search, and multiple-level ``undo''
features in commercial applications. In the early nineties,
Baker~\cite{Baker:1992:LLL,Baker:1992:NFT} argued for a systematic, first-class,
treatment of reversibility but intensive research in full-fledged reversible
models of computations and reversible programming languages was only sparked by
the discovery of deep connections between physics and
computation~\cite{Landauer:1961, PhysRevA.32.3266,Toffoli:1980,
  bennett1985fundamental,Frank:1999:REC:930275} and by the potential for
efficient quantum computation~\cite{springerlink:10.1007/BF02650179}.

The early developments of reversible programming languages started with a
conventional programming language, e.g., extended $\lambda$-calculi, and either
(i) extended the language with a history
mechanism~\cite{vanTonder:2004,Kluge:1999:SEMCD,lorenz,danos2004reversible} or
(ii) imposed constraints on the control flow constructs to make them
reversible~\cite{Yokoyama:2007:RPL:1244381.1244404}. More modern approaches
recognize that reversible programming languages require a fresh approach and
should be designed from first principles without the detour via conventional
irreversible languages~\cite{Yokoyama:2008:PRP, Mu:2004:ILRC,
  abramsky2005structural, DiPierro:2006:RCL:1166042.1166047}.

@@

In previous work, Carette, Bowman, James, and
Sabry~\cite{rc2011,James:2012:IE:2103656.2103667,Carette2016} introduced
the~$\Pi$ family of reversible languages based on type isomorphisms and
commutative semiring identities. The fragment without recursive types is
universal for reversible boolean
circuits~\citep{James:2012:IE:2103656.2103667}. With the addition of recursive
types and trace operators~\citep{Hasegawa:1997:RCS:645893.671607}, the variant
of $\Pi$ becomes a Turing-complete reversible
language~\citep{James:2012:IE:2103656.2103667,rc2011}. Before identifying a
``featherweight'' representative to use for the formal development in this
paper, we illustrate the main underlying ideas for the entire family of
languages using some examples.

@@

Although writing circuits using the raw syntax for combinators is tedious, the
examples illustrate the programming language nature of this family of $\Pi$
languages. In other work, one can find a compiler from a conventional functional
language to circuits~\citep{James:2012:IE:2103656.2103667}, a systematic
technique to translate abstract machines to $\Pi$~\citep{rc2012}, and a
Haskell-like surface language~\citep{theseus} which can ease writing
circuits. All that reinforces the claim that this is a practical approach to
reversible programming.

>>>>

\documentclass{entcs}
\usepackage{prentcsmacro}

\def\lastname{Carette, Chen, Choudhury, Rose, and Sabry}

@@

\begin{frontmatter}
\title{From Reversible Programs to \\ Univalent Universes and Back}
\author{Jacques Carette}
\address{McMaster University}
\author{Chao-Hong Chen}
\address{Indiana University}
\author{Vikraman Choudhury}
\author{Robert Rose}
\author{Amr Sabry}
\address{Indiana University}

@@

\end{frontmatter}

@@

The practice of programming languages is replete with \emph{ad hoc} instances of
reversible computations such as database transactions, mechanisms for data
provenance, checkpoints, stack and exception traces, logs, backups, rollback
recoveries, version control systems, reverse engineering, software transactional
memories, continuations, backtracking search, and multiple-level ``undo''
features in commercial applications. In the early nineties,
Baker~\cite{Baker:1992:LLL,Baker:1992:NFT} argued for a systematic, first-class,
treatment of reversibility but intensive research in full-fledged reversible
models of computations and reversible programming languages was only sparked by
the discovery of deep connections between physics and
computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275}
and by the potential for efficient quantum
computation~\cite{springerlink:10.1007/BF02650179}.

The early developments of reversible programming languages started with a
conventional programming language, e.g., extended $\lambda$-calculi, and either
(i) extended the language with a history
mechanism~\cite{vanTonder:2004,Kluge:1999:SEMCD,lorenz,danos2004reversible} or
(ii) imposed constraints on the control flow constructs to make them
reversible~\cite{Yokoyama:2007:RPL:1244381.1244404}. More modern approaches
recognize that reversible programming languages require a fresh approach and
should be designed from first principles without the detour via conventional
irreversible languages~\cite{Yokoyama:2008:PRP,Mu:2004:ILRC,abramsky2005structural,DiPierro:2006:RCL:1166042.1166047}.

@@

In previous work, Carette, Bowman, James, and
Sabry~\cite{rc2011,James:2012:IE:2103656.2103667,Carette2016} introduced
the~$\Pi$ family of reversible languages based on type isomorphisms and
commutative semiring identities. The fragment without recursive types is
universal for reversible boolean
circuits~\cite{James:2012:IE:2103656.2103667}. With the addition of recursive
types and trace operators~\cite{Hasegawa:1997:RCS:645893.671607}, the variant
of $\Pi$ becomes a Turing-complete reversible
language~\cite{James:2012:IE:2103656.2103667,rc2011}. Before identifying a
``featherweight'' representative to use for the formal development in this
paper, we illustrate the main underlying ideas for the entire family of
languages using some examples.

@@

Although writing circuits using the raw syntax for combinators is tedious, the
examples illustrate the programming language nature of this family of $\Pi$
languages. In other work, one can find a compiler from a conventional functional
language to circuits~\cite{James:2012:IE:2103656.2103667}, a systematic
technique to translate abstract machines to $\Pi$~\cite{rc2012}, and a
Haskell-like surface language~\cite{theseus} which can ease writing
circuits. All that reinforces the claim that this is a practical approach to
reversible programming.
